---
title: 'Shiny Exercise 2: The One Sample Bootstrap'
author: "CMSC 205, Winter 2017"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Possible solution to Friday's exercise

While there is no "right" way to create such an app, you definitely want to break down the task into bite-sized pieces. For example I

1. Got the ggplot working using "pretend" inputs, then tested
2. Made the name, start_year, and end_year input working one-by-one (but didn't connect to plot yet), then tested
3. Then synced `input$name`, `input$start_year`, and `input$end_year` into ggplot one-by-one, then tested



## Modularize code with reactive()

Sometimes you need to create a reactive object outside of a `renderSomething` chunk. For example, if you need to access that object from multiple `renderSomething` chunks.

In this situation, use the `reactive` function. 




## Task

Using the `sf2016_sample.csv` data set, create a shiny app to run a one-sample bootstrap for the mean. Specifically, a user should be able to

- specify the number of bootstrap resamples to draw with a slider, 
- select `Salaries` or `TotalCompensation` from a drop-down menu, and
- enter the confidence level for a percentile confidence interval. 

The results should include:

- a histogram of the original sample,
- a histogram of the bootstrap distribution for the mean,
- the mean of the bootstrap distribution along with the percentile confidence interval.

## Resources/Hints

1. Below is the one-sample bootstrap function we wrote in class on 2/8. Note that running the .Rmd file will make this function available.

```{r}
# x = the data vector
# stat = an R function to calculate the statistic of interest
# B = the number of bootstrap resamples
bootstrap1 <- function(x, stat, B, na.rm = TRUE) {
  n <- length(x)
  stats <- numeric(length = B) # Place to store the stats
  for(b in 1:B) {
    boot_sample <- sample(x, size = n, replace = TRUE)
    stats[b] <- stat(boot_sample, na.rm = TRUE)
  }
  stats
}
```

2. To select a column based on an input, use the `select` function in the `dplyr` package along with the `one_of` function with the input.

3. When specifying the aesthetics in `ggplot`, use `aes_string` instead of `aes` if you need to specify the name of a column as a string.

## Advice

This is a complicated Shiny app---there are a lot of moving parts. I recommend that you 

- Start with a "brain-dead" version and go end-to-end immediately.
- Then back and iterate and improve.

You get feedback much quicker this way!

## Build the App



Below is the Shiny template we discussed last class. Edit the template in order to create your app.

```{r eruptions, echo=FALSE}
inputPanel(
  selectInput("n_breaks", label = "Number of bins:",
              choices = c(10, 20, 35, 50), selected = 20),
  
  sliderInput("bw_adjust", label = "Bandwidth adjustment:",
              min = 0.2, max = 2, value = 1, step = 0.2)
)

renderPlot({
  hist(faithful$eruptions, probability = TRUE, breaks = as.numeric(input$n_breaks),
       xlab = "Duration (minutes)", main = "Geyser eruption duration")
  
  dens <- density(faithful$eruptions, adjust = input$bw_adjust)
  lines(dens, col = "blue")
})
```

