---
title: 'Shiny Exercise 2: The One Sample Bootstrap'
author: "CMSC 205, Winter 2017"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Modularize code with reactive()

Sometimes you need to create a reactive object outside of a `renderSomething` chunk. For example, if you need to access that object from multiple `renderSomething` chunks.

In this situation, use the `reactive` function. 


## Task

Using the `sf2016_sample.csv` data set, create a shiny app to run a one-sample bootstrap for the mean. Specifically, a user should be able to

- specify the number of bootstrap resamples to draw with a slider, 
- select `Salaries` or `TotalCompensation` from a drop-down menu, and
- enter the confidence level for a percentile confidence interval. 

The results should include:

- a histogram of the original sample,
- a histogram of the bootstrap distribution for the mean,
- the mean of the bootstrap distribution along with the percentile confidence interval.

## Resources/Hints

1. Below is the one-sample bootstrap function we wrote in class on 2/8. Note that running the .Rmd file will make this function available.

```{r}
# x = the data vector
# stat = an R function to calculate the statistic of interest
# B = the number of bootstrap resamples
bootstrap1 <- function(x, stat, B, na.rm = TRUE) {
  n <- length(x)
  stats <- numeric(length = B) # Place to store the stats
  for(b in 1:B) {
    boot_sample <- sample(x, size = n, replace = TRUE)
    stats[b] <- stat(boot_sample, na.rm = TRUE)
  }
  stats
}
```

2. To select a column based on an input, use the `select` function in the `dplyr` package along with the `one_of` function with the input.

3. When specifying the aesthetics in `ggplot`, use `aes_string` instead of `aes` if you need to specify the name of a colum as a string.

## Advice

This is a complicated Shiny app---there are a lot of moving parts. I recommend that you 

- Start with a "braindead" version and go end-to-end immediately.
- Then back and iterate and improve.

You get feedback much quicker this way!

## Build the App



Below is the Shiny template we discussed last class. Edit the template in order to create your app.

```{r eruptions, echo=FALSE}
# Load the data set
sf_sample <- read.csv("sf2016_sample.csv")

inputPanel(
  sliderInput("n_boot", label = "Number of bootstrap resamples:", 
              min = 500, max = 10000, value = 1000),
  
  selectInput("variable", label = "Variable:",
              choices = c("Salaries", "TotalCompensation")),
  
  numericInput("conf_level", label = "Confidence level", value = 0.95, min = 0.8, max = 0.99, step = 0.01)
)

data <- reactive(select(sf_sample, one_of(input$variable)))

renderPlot({
  ggplot(data = data(), aes_string(x = input$variable)) +
    geom_histogram(color = "black") +
    labs(x = input$variable, title = paste("Distribution of", input$variable))
})

means <- reactive({bootstrap1(data()[[1]], stat = mean, 
                             B = input$n_boot, na.rm = TRUE)})

renderPlot({
  ggplot(data.frame(means = means()), aes(x = means)) +
    geom_histogram(color = "black") +
    labs(x = paste("Mean", input$variable), title = "Bootstrap Distribution")
})

renderPrint({
  alpha <- 1 - input$conf_level
  c(mean = mean(means()), quantile(means(), alpha/2), 
    quantile(means(), 1 - alpha/2))
})
```

