---
title: "Writing Functions"
output: html_notebook
---

## Overview 

### Basic anatomy of a function

```{r eval=FALSE}
￼my_fun <- function(arg1, arg2) {
  body 
}
```

### Example: Circumference of a circle

The circumference of a circle is $2\pi r$ for a given radius, r. Let's write a function called `circumference` that performs this calculation for any given radius.

```{r}
# Define circumference() here
circumference <- function(r) {
  2 * pi * r
}
circumference(2)
```


### Example: Dealing cards

The below code defines a standard 52-card deck.

```{r warning=FALSE}
deck <- as.vector(outer(c("A",2:10,"J","Q","K"), c("\u2660","\u2665","\u2666", "\u2663"), paste0))
deck
```

Let's write a function called `deal` that deals an `n` card hand from a deck, `x`.

```{r}
# Define deal() here
deal <- function(x, n) {
  sample(x, size = n)
}
```

Run the below code to deal a 5-card hand and a 13-card hand:

```{r}
deal(deck, n = 5)

deal(deck, n = 13)
```


Here, we might consider setting a default value for 

```{r}
# Alter deal() so that n = 5 by default
deal <- function(x, n = 5) {
  sample(x, size = n)
}
```

What happens if we run the below code?

```{r}
deal(deck)
```


* * *

## Your turn: Gambler's ruin

Below is the code we wrote to simulate one trial in the gambler's ruin problem. Write a function called `gamblers_ruin` that takes `N` (total wealth), `p` (probability that A wins), and `i` (A's initial wealth) as inputs and returns the number of bets until one player is ruined.

```{r}
N <- 20   # set total wealth
p <- 0.5  # set prob. A wins a bet
i <- 10   # A's initial wealth

gamblers_ruin <- function(N, p, i) {
  bets <- 0 # initializing no. of bets
  
  while(i > 0 & (N - i) > 0) {
    bets <- bets + 1
    outcome <- rbinom(1, n = 1, p = p)
    if (outcome == 1) {
      i <- i + 1
    } else {
      i <- i - 1
    }
  }
  
  bets
}
```

```{r}
# Test gamblers_ruin() here
N <- 20   # set total wealth
p <- 0.5  # set prob. A wins a bet
i <- 10   # A's initial wealth

gamblers_ruin(N = 100, p= 0.8, i = 10)
```

Writing a function to perform the simulation allows us to easily replicate it's output many times. In this situation, we can use the `replicate` function to quickly run our function 1,000 times.

```{r}
our_sim <- replicate(n = 1000, expr = gamblers_ruin(N = 20, p = .5, i = 10))
head(our_sim)
```

Then we can easily calculate the expected number of bets required for one player to be ruined.

```{r}
mean(our_sim)
```

We can also plot the simulation results

```{r message = FALSE}
library(tidyverse)
ggplot(data = data.frame(bets = our_sim), aes(x = bets)) +
  geom_histogram(color = "gray20", binwidth = 20)
```



## Your turn: One-sample bootstrap for the median

Below is the code we wrote last class to carryout the one-sample bootstrap for the sample median based on `our_sample`. Write a function called `bootstrap_median` that performs this bootstrap and returns the medians. Take a vector, `x`, and the number of bootstrap resamples, `B`, as inputs.

```{r message=FALSE}
# Load the libraries
library(tidyverse)
library(readr)

# Read in the data sets
sf2016 <- read_csv("https://github.com/cmsc205/notes/raw/master/day11/sf2016.csv")

# Take a random sample from the population
our_sample <- sf2016 %>%
  sample_n(size = 500)

# Below is the code for bootstrapping the sample median
B <- 5000 # No. of bootstrap statsistics desired
n <- nrow(our_sample)
medians <- numeric(length = B) # Place to store the stats
for(b in 1:B) {
  boot_sample <- sample(our_sample$TotalCompensation, 
                        size = n, 
                        replace = TRUE)
  medians[b] <- median(boot_sample)
}
head(medians)
```

```{r}
# Define bootstrap_median() here
bootstrap_median <- function(x, B = 5000) {
  n <- length(x)
  medians <- numeric(length = B) # Place to store the stats
  for(b in 1:B) {
    boot_sample <- sample(x, size = n, replace = TRUE)
    medians[b] <- median(boot_sample)
  }
  medians
}
```

```{r}
# Let's test the function
med1 <- bootstrap_median(our_sample$TotalCompensation)
```



## Your turn: One-sample bootstrap for the mean

Once you have written `bootstrap_median` it's easy to extend what you have learned to other one-sample bootstraps.

Write a function called `bootstrap_mean` that performs this bootstrap and returns the means Take a vector, `x`, and the number of bootstrap resamples, `B`, as inputs.

```{r}
# Define bootstrap_mean() here
bootstrap_mean <- function(x, B = 5000) {
  n <- length(x)
  means <- numeric(length = B) # Place to store the stats
  for(b in 1:B) {
    boot_sample <- sample(x, size = n, replace = TRUE)
    means[b] <- mean(boot_sample)
  }
  means
}
```

Now that you have written two one-sample bootstrap functions, compare your code. How many of the commands are the same? Did it feel like you were simply copying and pasting?


* * *

## Using a function as an argument

The only differences between `bootstrap_median` and `bootstrap_mean` are the statistics calculate from each sample, which resulted in a lot of repetition between the two functions. An alternative to writing different bootstrapping functions for different statistics is to specify the function of interest as an argument to a more general function.

Let's write a single `bootstrap1` function that conducts a one-sample bootstrap on a vector, `x`. We'll use the `stat` argument to specify the statistic of interest, and `B` to denote the number of bootstrap samples desired.

```{r}
# Define the bootstrap1() function here
bootstrap1 <- function(x, stat, B, na.rm = TRUE) {
  n <- length(x)
  stats <- numeric(length = B) # Place to store the stats
  for(b in 1:B) {
    boot_sample <- sample(x, size = n, replace = TRUE)
    stats[b] <- stat(boot_sample, na.rm = TRUE)
  }
  stats
}
```


Let's see how it works on the `our_sample` data.

```{r}
boot_medians <- bootstrap1(our_sample$TotalCompensation, 
                           stat = median, B = 5000)

boot_means <- bootstrap1(our_sample$TotalCompensation, 
                         stat = mean, B = 5000)
```


* * *

## Assignment

Define a function implementing the two-sample bootstrap, which I have defined below. Name the function `bootstrap2`, and use the following argument names:

* `x` -- vector containing the sample from group 1
* `y` -- vector containing the sample from group 2
* `stat` -- function comparing
* `B` -- number of bootstrap resamples
* `ratio` -- a logical value. If FALSE then statistics for two samples are combined using statistic1 - statistic2 (the statistics from the two samples). If TRUE, it uses statistic1 / statistic2.


Use this function to create a bootstrap distribution for the difference in average salaries between `Community Health` and `Public Protection` jobs. (These categories are defined in the `OrganizationGroup` column.) Display the bootstrap distribution using a histogram.

#### Two-sample bootstrap algorithm

Given independent samples of sizes m and n from two groups, 

1. Draw a resample of size m, with replacement, from the first sample and a separate resample of size n, with replacement, from the second sample. Compute a statistic that compares the two groups, such as the difference between the two sample means. 
2. Repeat this resampling process many times, say 10,000. 
3. Construct the bootstrap distribution of the statistic. 
