---
title: "Our First App"
author: "CMSC 205, Winter 2017"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(babynames)
```


While there is no "right" way to create such an app, you definitely want to break down the task into bite-sized pieces. For example I

1. Got the ggplot working using "pretend" inputs, then tested
2. Made the name, start_year, and end_year input working one-by-one (but didn't connect to plot yet), then tested
3. Then synced `input$name`, `input$start_year`, and `input$end_year` into ggplot one-by-one, then tested


```{r babynames, echo=FALSE}
inputPanel(
  # Input name
  textInput("name", label = "Enter Name:", value="Jane"),
  # Input start year
  numericInput("start_year", label = "Start Year:", value = 1880, min = 1880, max = 2014, step = 1),
  # Input end year
  numericInput("end_year", label = "End Year:", value = 2014, min = 1880, max = 2014, step = 1)
)

renderPlot({
  # Generate Plots
  names <- babynames %>%
    filter(name == input$name, year >= input$start_year, year <= input$end_year )
  ggplot(data=names, aes(x=year, y=prop, col=sex)) + 
    geom_line() +
    scale_color_manual(values = c("#e41a1c", "#377eb8"))
})
```
