---
title: "Common plots in ggplot2"
output: html_notebook
---

The type of plot created depends on a few factors:

- the type of variable(s) being displayed,
- and the number of variables being displayed.

Today we'll learn more about creating specific types of plots using the `ggplot2` package.

To begin, load the `tidyverse`:

```{r message=FALSE}
library(tidyverse)
```

Further, to ground our discussion, let's consider the `diamonds` data set that is loaded with the `tidyverse`.

```{r}
head(diamonds)
```

**Question.** How how many observations are in the diamonds data set?

For information about the variable definitions, see the help file using `?diamonds` in the console.


# Univariate graphics

### Histograms

**Use:** exploring the distribution of a quantitative variable

**geometry:** `geom_histogram`

**aesthetics required**: `x` (the variable to be displayed)


```{r}
# Histogram of carat

```

**Remember to check the binwidth!**


### Density plots

**Use:** exploring the distribution of a quantitative variable

**geometry:** `geom_density`

**aesthetics required**: `x` (the variable to be displayed)


```{r}
# Density plot of price

```


### Bar charts

**Use:** exploring the distribution of a categorical variable

**geometry:** `geom_bar`

**aesthetics required**: `x` (the variable to be displayed)


```{r}
# Bar chart of clarity

```


* * * 

# Bivariate graphics

### Scatterplots

**Use:** exploring the relationship between two quantitative variables

**geometry:** `geom_point`

**aesthetics required**: `x`, `y`

```{r}
# Let's see how carat and price are related
```

But that's pretty hard to read because we're plotting so many points!

```{r}
# Set the alpha level to help with overplotting
```

We can also add a smoother to reveal the trend in the scatterplot

```{r}
# Add a smoother to the plot above
```




### Segmented bar chart

**Use:** exploring the relationship between two categorical variables

**geometry:** `geom_bar`

**aesthetics required**: `x`, `fill`

```{r}
# create a segmented bar chart of  clarity by color
```


It's also interesting to set the `position`.

```{r}
# create a segmented bar chart of  clarity by color, setting position = "fill"
```

```{r}
# create a segmented bar chart of  clarity by color, setting position = "dodge"
```

### Side-by-side boxplots


**Use:** exploring the relationship between a quantitative and categorical variable

**geometry:** `geom_boxplot`

**aesthetics required**: `x` (categorical), `y` (quantitative)

```{r}
# create boxplots of price by cut
```

```{r}
# Let's try adding the points on top of the boxplots
```

```{r}
# Let's try again with geom_jitter
# we'll need to eliminate the outliers in the boxplots by setting
#   outlier.size=0
```





* * * 

# Adding in more variables

To add additional variables we can use **aesthetics** such as

- size
- color
- fill
- shape

```{r}
# Let's look at whether adding color for cut tells us anything about
# the relationship between carat and price
```

```{r}
# Would shape be a better aesthetic to use?
```

```{r}
# Would be using both color and shape better?
```


Alternatively, we can use **facetting**.

```{r}
# Let's facet the scatterplot of price vs. carat by color
```

```{r}
# Let's facet the scatterplot of price vs. carat by color AND clarity
```



* * * 

# Exercise

In this exercise, we investigate home prices in Saratoga County, New York. The data set `saratoga.csv` contains the prices of 1,728 homes. 

To begin, load the data set.

```{r}
# SOLUTION
```


1. Describe the distribution of `Price`.

```{r}
# SOLUTION
```


2. Are houses with fireplaces generally more expensive than houses without?

```{r}
# SOLUTION
```

3. Describe the relationship between Price and Living Area.

```{r}
# SOLUTION
```

4. Update the plot you created in part 3 to include whether information about whether the house has a fireplace. Summarize your findings.

```{r}
# SOLUTION
```


