---
title: "4/26 Solutions"
author: "CMSC 205, Spring 2017"
output: html_notebook
---

## Tidying french fries

```{r Loading ff}
ff <- read.csv("https://raw.githubusercontent.com/cmsc205/data/master/frenchfries.csv")

library(tidyverse)
glimpse(ff)
```

# Goals
- Create a scale column (potato, buttery, etc.)
- Create a rating column to store the values from the old columns
- Create boxplots of ratings by scale 

```{r ff tidy}
tidy_ff <- ff %>%
  gather(key = "scale", value = "rating", potato, buttery, grassy, rancid, painty)
glimpse(tidy_ff)
```

```{r boxplots}
ggplot(tidy_ff, aes(x = scale, y = rating)) +
  geom_boxplot() +
  facet_wrap(~time, ncol = 5)
```

```{r}
ggplot(tidy_ff, aes(x = factor(time), y = rating)) +
  geom_boxplot() +
  facet_wrap(~scale)
```


```{r}
ggplot(tidy_ff, aes(x = time, y = rating, color = scale)) +
  # geom_point() +
  geom_smooth(se = FALSE) 
  # facet_wrap(~scale)
```



## Example of separating

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/")
```

## Tidying the polls.csv data set

```{r}
polls <- read.csv("https://github.com/cmsc205/notes/raw/master/day13/polls.csv")
```

We discussed this solution in class, so I will omit the explanation of why the data are not tidy.

```{r tidying}
polls %>%
  separate(Date, into = c("start", "end"), "-") %>%
  separate(Sample, into = c("size", "population")) %>%
  gather(key = "candidate", value = "percent", Clinton..D.:Stein..G.) %>%
  separate(candidate, into = c("candidate", "party"), sep = "[..]") %>%
  mutate(MoE = ifelse(MoE == "--", NA, MoE))
```


## Tidying the under5mortality.csv data set

```{r}
u5_mort <- read.csv("https://github.com/cmsc205/notes/raw/master/day13/under5mortality.csv")
glimpse(u5_mort)
```

Notice that years are specified in the column names, but the year is a variable and should have it's own column. Also, remember that rows should correspond to the observational unit. Countries in a specific year are observed, so each country in each year should have a separate row.

```{r}
u5_mort %>%
  rename(country = Under.five.mortality) %>%
  gather(key = "year", value = "u5_mortality", -country) %>%
  mutate(year = parse_number(year))
```


## Tidying the mlb2016.csv data set

```{r}
mlb2016 <- read.csv("https://github.com/cmsc205/notes/raw/master/day13/mlb2016.csv")
glimpse(mlb2016)
```

In this data set the `Years` column contains three variables: the length of a players contract, and the start and end dates. Additionally, the salary columns contain units (`$`) and commas, forcing them to be factors.

```{r}
mlb2016 %>%
  separate(Years, into = c("contract_length", "start_year", "end_year")) %>%
  mutate(contract_length = as.integer(contract_length),
         end_year = paste(20, end_year, sep = ""),
         Salary = parse_number(Salary),
         Total.Value = parse_number(Total.Value),
         Avg.Annual = parse_number(Avg.Annual))
```

