---
title: "Introduction to dplyr"
output: html_notebook
---

## Overview

Data manipulation is central to data analysis and is often the most time consuming portion of an analysis. The `dplyr` package contains a suite of functions to make data manipulation easier. The core functions of the `dplyr` package can be thought of as verbs for data manipulation.


Verb(s)               | Meaning
------------ | -----------------------------------------------------
`filter`     | pick specific observations (i.e. specific rows)
`arrange`    | reorder the rows
`select`     | pick variables by their names (i.e. specific columns)
`mutate`     | add new calculated columns to a data frame
`summarize`  | aggregate many rows into a single row 


In this example you will explore how to use each of these functions.

To begin, load the `tidyverse` (which loads `dplyr`) and the data set

```{r message=FALSE, warning=FALSE}
colleges <- read.csv("https://raw.githubusercontent.com/cmsc205/data/master/colleges2015.csv")
library(tidyverse)
```


## Data 

The file `college2015.csv` contains information on predominantly bachelor's-degree granting institutions from 2015 that might be of interest to a college applicant. A brief summary of each column is given below:


Variable        | Description
--------------- | -----------
`unitid`        | A unique ID number for each school
`college`       | School name
`type`          | School type: public or private
`city`          | City
`state`         | State abbreviation
`region`        | Region of the U.S.
`admissionRate` | Proportion of undergraduate applicants admitted
`ACTmath`       | Median ACT math score
`ACTenglish`    | Median ACT English score
`undergrads`    | Undergraduate enrollment
`cost`          | Total cost of attendance
`gradRate`      | Proportion of students graduating within six years
`FYretention`   | Proportion of first year students returning for a second year
`fedloan`       | Proportion of students with federal student loans
`debt`          | Median principal of student loans entering repayment



## Questions

1) Create a new data frame that only contains the columns `college`, `city`, `state`, `undergrads`, and `cost`.

```{r}
sub_colleges <- select(colleges, college, city, state, undergrads, cost)
head(sub_colleges)
```

2) What is the most expensive school? 

```{r}
sub_colleges <- arrange(sub_colleges, desc(cost))
head(sub_colleges)
```

3) What school has the lowest admissions rate? 

```{r}
head(arrange(colleges, admissionRate))
```

4) What school has the least expensive school in Wisconsin?

```{r}
wi <- filter(colleges, state == "WI")
head(arrange(wi, cost))
```

5) What school has the lowest admissions rate in the Great Lakes or Plains regions?

```{r}
glp <- filter(colleges, region == "Great Lakes" | region == "Plains")
glp <- filter(colleges, region %in% c("Great Lakes", "Plains"))
head(arrange(glp, admissionRate))
```

5) How many private colleges are there in Maryland with under 5000 undergraduates?

```{r}
md <- filter(colleges, state == "MD" & type == "private" & undergrads < 5000)
summarize(md, N = n())
```


6) Create a column giving the cumulative average cost of attendance, assuming that students finish in four years and that costs increase 3% per year. Name this new column `total.avg.cost4`.

```{r}
sub_colleges <- mutate(sub_colleges, total.avg.cost4 = cost + 1.03 * cost + 1.03^2 * cost + 1.03^3 * cost)
head(sub_colleges)
```




## Additional Resources

* [RStudio's data wrangling cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) provides a nice summary of the functions in the `dplyr` package, including those covered in this tutorial.

* The [introductory vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html) to `dplyr` provides an example of wrangling a data set consisting of 336,776 flights that departed from New York City in 2013.

* Roger Peng's [video overview](https://www.youtube.com/watch?v=aywFompr1F4&feature=youtu.be) of the `dplyr` package.