---
title: "R Programming Structures: Loops and if-else statements"
output: html_document
---

## Setup 

First, we need to load the necessary packages.

```{r message=FALSE}
library(stringr)
library(rvest)
```

# Motivating Example: Web Scraping

## Quick note on HTML

Web pages are written in HTML (Hyper Text Markup Language) which uses **tags** to describe different aspects of document content. 

* a heading in a document is indicated by `<h1>My Title</h1>` 
* a paragraph is indicated by  `<p>A paragraph of content...</p>`
* a table is indicated by `<table> ... </table>`

## Example

* [Box Office Mojo](http://www.boxofficemojo.com) gives statistics on box office earnings of movies. 

* Web site also maintains lists of yearly and all time record holders.

* We will look at the movies in the top 100 in all time movie worldwide grosses in box office receipts. 

* In particular, we will scrape the data from [Box Office Mojo: All Time Box Office](http://www.boxofficemojo.com/alltime/world/?pagenum=1). 

* The dollar amounts are in millions of dollars and the years marked with "^" indicate that the movie had multiple releases.


First, use `read_html` to read in the webpage

```{r}
movie_parse <- read_html("http://www.boxofficemojo.com/alltime/world/?pagenum=1")
```

Then, use `html_nodes` to grab the tables

```{r}
movie_tables <- html_nodes(movie_parse, "table")
movie_tables
```

Next, extract the desired table

```{r}
movies <- html_table(movie_tables[[3]], header = TRUE, fill = TRUE)
head(movies)
```


#### Question: Is the data set clean?

* * *

## Loops

We can use loops to handle this type of repetitive task. There are three types of loops:

1. `for` loop
2. `while` loop
3. `repeat` loop


### `for` loop

The `for()` statement allows one to specify that a certain operation should be repeated a fixed number of times.


##### Syntax

```{r eval=FALSE}
for(name in vector) {
  Commands
}
```


A *very* basic example

```{r}
for(i in 1:5) {
  print(i)
}
```


A second *very* basic example

```{r}
x <- c(5, 12, 13)
for(n in x) {
  print(n^2)
}
```


Now, let's use a `for` loop to clean our data set.

```{r}
problem_cols <- 4:9
for(col in problem_cols) {
  movies[,col] <- str_replace_all(movies[,col], pattern = "[^0-9.]", 
                                  replacement = "") %>%
    as.numeric()
}
```


<!-- We could also use a `for` loop to implement the sieve of Eratosthenes, the oldest known systematic method for listing prime numbers up to a given value n. -->

<!-- Idea:  -->

<!-- - Begin with a vector of numbers from 2 to n -->
<!-- - Beginning with 2, eliminate all multiples of 2 which are larger than 2 -->
<!-- - Move to the next number remaining in the vector, in this case, 3. Now, remove all multiples of 3 which are larger than 3.  -->
<!-- - Proceed through all remaining entries of the vector in this way. -->


<!-- ```{r} -->
<!-- n <- 50 -->
<!-- sieve <- seq(2, n) -->
<!-- primes <- c() -->
<!-- for(i in sieve) { -->
<!--   if(any(sieve == i)) { -->
<!--     primes <- c(primes, i) -->
<!--     sieve <- c(sieve[(sieve %% i) != 0], i) -->
<!--   } -->
<!-- } -->

<!-- primes -->
<!-- ``` -->



## `if` statement

The `if()` statement allows us to control which statements are executed.


##### Syntax

```{r eval=FALSE}
if(condition) {
  commands when TRUE
}
```


```{r eval=FALSE}
if(condition) {
  commands when TRUE
} else {
  commands when FALSE
}
```


A very basic example

```{r}
x <- 5
if (x > 5) {x <- x + 1}

x <- 6
if (x > 5) {x <- x + 1}
```


```{r}
x <- 5
if (x > 5) {
  x <- x + 1
} else {
  x <- x - 1
}
```



### `while` loop

The `while()` statement allows us to repeat a process as long as a condition holds.

##### Syntax

```{r eval=FALSE}
while (condition) {
  Commands
}
```

A very basic example

```{r}
x <- 0
while(x <= 5) {
  print(x)
  x <- x + 1
}
```


`while` loops are also very useful for simulation when you don't know how many trials you will need.


#### Example: Gambler's ruin

Two gamblers, A and B, make a sequence of $1 bets. In each bet, gambler A has probability p of winning, and gambler B has probability q = 1 − p of winning. Gambler A starts with i dollars and gambler B starts with N − i dollars; the total wealth between the two remains constant since every time A loses a dollar, the dollar goes to B, and vice versa. The game ends when either A or B is ruined, i.e., when the random walk reaches 0 or N. What is the probability that A wins the game (walking away with all the money)?

Note: `rbinom(1, n = 1, p = 0.5)` will simulate a coin flip where `1` denotes heads and `p` is the probability of heads.

