---
title: "Introduction to R"
output:
  html_notebook: default
  html_document: default
---

## Overview of R Markdown Notebooks

In this course we will often work with R markdown notebooks in class. When you execute code within the notebook, the results appear beneath the code. An alternative to the R markdown notebook is an R markdown file.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd/Cntrl+Shift+Enter*. 

```{r}
1:10
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd/Cntrl+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).


## Your First R session

While you are learning the R language, remember that you are learning a new language; thus, we will start rather simply with small analysis tasks and build up to more complicated tasks. Also, you will not remember everything immediately---that's OK, it's a natural part of learning a language! 

### Installing and loading R packages

R does not enable all of its functionality when you open it. To enable additional functionality we need to load *R packages*. In this class we will often use the `tidyverse` R package to enable better tools for plotting, data manipulation, etc. Below is an example of installation and loading:

```{r message=FALSE}
# You only need to install a package once
install.packages("tidyverse")

# You will need to load the package in each R markdown notebook
library(tidyverse)
```


### Loading data and assignment

Data science focuses on **data**, so we need to start with a data set. There are multiple ways to load data sets (and we'll learn more about them later), but we often need to load a `.csv` (comma seperated value) file.

Today, we'll look at a data set containing over 5,000 movies scraped from IMDB.

```{r}
# Load the data and assign it a name
movies <- read.csv("https://github.com/cmsc205/data/raw/master/movie_metadata.csv", as.is = TRUE)
```


### Data frames

The `movies` object is our first example of a data frame, which is essentially a list of vectors. We can get a first glimpse of our data set in a few ways:

```{r}
# Printing the first 6 rows
# Note that missing values are denoted by NA

```

```{r}
# Printing the last 6 rows

```

```{r}
# Looking at the number of rows and columns

```

```{r}
# Looking at the structure

```

```{r}
# Getting a glimpse of the structure

```


```{r}
# looking at the summary

```


#### Variable types

The `str` function revealed the structure of the `movies` data frame, which included the number of rows and columns, as well as a glimpse at each **variable** (i.e. column) and the first few cases.

In R, there are a few basic variable types (modes):

* Integer 
* Numeric (double, real)
* Character  
* Factor 
* Logical

There are other modes, but for now that this list will suffice.

#### Operators and expressions

As a data scientist you will want to **do** something with your data. So we need start adding things to your toolkit.

To begin, we can think of R as an over-powered calculator

```{r}
# Addition


# Subtraction

```

If you don't want to lose a quantity, then you should assign it a name.

```{r}
# Assign a name


# Multiplication


# Division


# Exponentiation

```

R obeys the familiar order of operations

```{r}
# Order of operations
1 + 2 * 3 * 4 * 5 / 6^3 + 7 + 8

1 + 2 * ((3 * 4 * 5 / 6)^3) + 7 + 8
```

But this is a data science class, so we need to bring back the data! For example, we may want to calculate the net profit in millions

```{r}

```



#### Syntax: Function application

We have already used functions in R---for example, `summary` is a function---so you have seen the syntax, but we need to make sure we understand what's happening.

The basic pattern is

```{r eval=FALSE}
object_name <- function_name(arguments)
```

Since we often have a data set of interest, many of our functions will take the form

```{r eval=FALSE}
object_name <- function_name(data, arguments)
```

