---
title: "Customizing Plots"
output: html_notebook
---

We now know how to create many plot types using `ggplot2`, but often you want to customize your plots. Today, we'll discuss common tools for customizing our plots.

To begin, make sure that you have the `tidyverse` loaded.

```{r message=FALSE}
library(tidyverse)
```

Also, let's start with our plot from the exercise

```{r}
scorecard <- read_csv("https://github.com/cmsc205/data/raw/master/ScorecardSmallNarrow.csv")
```

```{r}
# Code for your solution to the warm up exercise
# Let's assign the plot ot the object p
```


# Adding titles, axis labels, and legend labels


Good labels are critical for making your plots accessible to a wider audience. It's important to ensure axis and legend labels display the full variable name (or a logical abbreviaition), but this does not always happen by default. 

To add/alter these annotations we simply add a new layer using the `labs` function.


#### Adding a title

```{r}
# Code for adding a title to a plot
```

#### Adding a subtitle

```{r}
# Code for adding a subtitle to a plot
```

#### Changing the axis labels

```{r}
# Code to change axis labels
```


#### Changing the legend labels

```{r}
# Code to change legend labels
```


# Ordering factors

Sometimes we need to order the levels of a factor to make a plot meaningful.

To examine this, let's look at a new data set which contains informatio on the 10 most popular baby names in 2014.

```{r}
babynames <- read_csv("https://github.com/cmsc205/data/raw/master/top10babynames2014.csv")
```

```{r}
# Let's work with this bar chart of the top 10 baby names in 2014
babynames %>%
  ggplot(aes(x = name, y = n, fill = sex)) +
  geom_bar(stat = "identity")
```

To do this with respect to another variable we can use the `reorder` function. 

```{r}
# Reordering the names by popularity in the bar chart
```



# Choosing color palettes

`ggplot2` does not choose the nicest looking palettes, and the palettes are typically not colorblind friendly. Luckily, it's quite easy to change the color palettes you use.

#### Manual adjustments

To manually adjust the colors we can use the functions `scale_color_manual` or `scale_fill_manual`

```{r}
# Code to manually adjust the plot colors
```


#### Using RColorBrewer

You can let the RColorBrewer package choose the colors for you using the `scale_color_brewer` and `scale_fill_brewer` functions. 

```{r}
# Code to use the scale_fill_brewer function
```


You can see all of the possible palettes in the package by running 

```{r}
RColorBrewer::display.brewer.all()
```


#### Using viridis

If you install the viridis package (this might not be on the server), then you can use the `scale_color_viridis` and `scale_fill_viridis` functions.

```{r}
# Example of how to use the viridis color scheme with a discrete variable
```

#### A colorblind friendly palette

The `ggthemes` package contains a colorblind-friendly palette, which can be adopted using the `scale_color_colorblind` and `scale_fill_colorblind` functions.


```{r}
# Using the colorblind friendly palette in ggthemes
```




# Choosing themes

By default `ggplot2` uses a grey background and guidlines on a plot; however, we don't always want our plots to appear this way. To change this the appearance, we need to work with themes.

The `ggthemes` package offers many additional [choices](https://github.com/jrnold/ggthemes). 

We can use an existing theme just like adding a layer.


```{r}
# Let's check out the minimal theme first to remove many features
```


```{r}
# Let's pick another theme from the list
```


# Customizing themes

Sometimes the `ggthemes` package won't do exactly what you want so you will have to dig into `ggplot2` a bit deeper and create your own theme.

See `?theme` for a list of all of the components that you can modify.


```{r}
# Let's really make a bare bones bar chart...
```

