---
title: "Intro to Text Manipulation in R"
output: html_document
---

## Set up for today

Install and load the `stringr` package. Remember that we **do not install packages in our R markdown file**, we only load them.

```{r}
library(stringr)
```


## Definition

A **(character) string** is any finite sequence of characters (i.e., letters, numerals, symbols and punctuation marks).


## Basic commands

In the `base` package (loaded by default), there are ways to manipulate character strings, most of which we won't discuss, but we do need to discuss a few:

`nchar` determines the length of a string.

```{r}
lu <- "Lawrence University"
nchar(lu)
```


`paste` allows us to easily create strings from variables/vectors.

```{r}
building <- "Briggs"
room <- "421"
begin_time <- "9:50 a.m."
end_time <- "11:00 a.m."
days <- "MWF"
class <- "CMSC 205"
paste(class, "meets from", begin_time, "to", end_time, days, "in", building, room, sep=" ")
```


`paste` also works with vectors

```{r}
paste(letters, 1:26)
paste(letters, 1:26, sep = "")
paste(letters, 1:26, sep = "-")
```


You can also make everything either lowercase or uppercase using `tolower` and `toupper`, respectively.

```{r}
text <- "AppleTon, Wi"
tolower(text)
toupper(text)
```



## stringr Package

Handling text can be complicated, but is often necessary. The `stringr` package eases the level of complication for many commonly used manipulations (i.e. it eliminates options that you often don't need to worry about).


## Extracting substrings

The `str_sub` command extracts substrings from a string given the starting and ending position.

Example with a single character string

```{r}
lu <- "Lawrence University"
str_sub(lu, 1, 8)
str_sub(lu, 1, nchar(lu))
str_sub(lu, c(1, 10), c(8, 19))
```

What about a vector of strings?

```{r}
fruits <- c("apple", "pineapple", "Pear", "orange", "peach", "banana")
str_sub(fruits, 2, 4)
str_sub(fruits, 2, 2)
str_sub(fruits, 4, 4)
str_sub(fruits, c(2,4), c(2,4))
str_sub(fruits, c(2,4,2,4,2,4), c(2,4,2,4,2,4))
```


## Padding strings

Sometimes, you want to add character(s) to the beginning or end of a string

```{r}
nums <- 1:15
as.character(nums)
str_pad(nums, 2, pad ="0")
str_pad(nums, 3, pad ="0")
```


## Detecting strings 

The `str_detect` command checks to see if **any** instance of a pattern occurs in a string.

```{r}
fruits <- c("apple", "pineapple", "Pear", "orange", "peach", "banana")
str_detect(fruits, pattern = "p")
```


## Matching strings

```{r}
strings <- c(" 219 733 8965", "329-293-8753 ", "banana", "595 794 7569",
             "387 287 6718", "apple", "233.398.9187  ", "482 952 3315",
             "239 923 8115 and 842 566 4692", "Work: 579-499-7527",
             "$1000", "Home: 543.355.3679")
str_match(string=strings, pattern="233")
str_match(string=strings, pattern="Work")

```


## Locate the position of a string

To locate the position of a pattern within a string, use `str_locate`.


```{r}
sentence <- "Peter Piper picked a peck of pickled peppers."
str_locate(sentence, pattern="pick")
str_locate_all(sentence, pattern="pick")
```


```{r}
fruits
str_locate(fruits, "an")
str_locate_all(fruits,"an")
output <- str_locate_all(fruits,"an")
output[[6]]
```


## String replacing

Use `str_replace` to replace a character string with 

```{r}
rhyme <- "Peter Piper picked a peck of pickled peppers; A peck of pickled peppers Peter Piper picked; If Peter Piper picked a peck of pickled peppers, Where's the peck of pickled peppers Peter Piper picked?"

str_replace(string=rhyme, pattern = "Peter Piper", replacement = "John Doe")

str_replace_all(string=rhyme, pattern = "Peter Piper", replacement = "John Doe")
```


## String splitting

To split a string on based on a specific character, use `str_split`.

```{r}
str_split(rhyme, pattern = ";")
str_split(rhyme, pattern = "; ")
```



## Common pitfalls

There are a number of meta characters that we cannot use directly in the above functions

`.  ^  \  $  ?  *  +  [  ]  (  )  {  }  |`

If you actually want to use these symbols literally, then in R, you precede them by a double slash (or surround them with brackets).


```{r}
tuition <- "$45,801"
str_replace(tuition, pattern = "$", replacement = "")
str_replace(tuition, pattern = "\\$", replacement = "")
str_replace(tuition, pattern = "[$]", replacement = "")

string <- "Hi["
str_replace(string, pattern = "\\[", replacement = "")
```



## Introduction to regular expressions

Sometimes the patterns we wish to detect, extract, replace, etc. are complex and we need to turn to **regular expressions**.

Regular expressions (regexps) are very terse and can be quite confusing, but sometimes they are necessary.

**Motivation:** We wish to anonymize phone numbers. 

```{r}
a1 <- "Home: 507-645-5489"
a2 <- "Cell: 219.917.9871"
a3 <- "My work phone is 507-202-2332"
a4 <- "I don't have a phone"
info <- c(a1, a2, a3, a4)
info
```


We need to build up to this solution slowly...


Question: How can we detect **any** digit?

```{r}
str_view(a1, "\\d")
str_view_all(a1, "\\d")

str_view(a1, "[0-9]")
str_view_all(a1, "[0-9]")
```


Question: How can we detect three digits in a row?

```{r}
str_view(a1, "[0-9]{3}")
```


Question: How can we detect either `.` or `-`?

```{r}
str_view(a1, "[.-]")
str_view(a2, "[.-]")
```


OK, let's put this together and try to extract the phone numbers from `info`.

```{r}
str_extract(info, pattern = "[2-9][0-9]{2}[.-][0-9]{3}[.-][0-9]{4}")
```

Now, let's replace the numbers with an anonymized `"XXX-XXX-XXXX"`.


```{r}
str_replace_all(info, pattern = "[2-9][0-9]{2}[.-][0-9]{3}[.-][0-9]{4}", replacement = "XXX-XXX-XXXX")
```

